\bf A class of exponential family measurement error models for singe-cell CRISPR screens

### 2022-01-04 text/GLM-EIV/jasa_v2.tex
fix reproducibility

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in a simplified Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model \parencite{carroll} to responses and noisy predictors that are exponential family-distributed and potentially impacted by the same set of confounding variables. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding. We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.

@@

\section{Background and analysis challenges}
\iffalse
\textcolor{red}{=====================================}
\textcolor{blue}{I don't think this section works as is.  I think the whole section should be relocated and revised, but let me say what you might do, if you keep it here. First, as much as I like the eSVD paragraph, I would argue that we aren't the only ones using negative binomial models for CRISPR screen analysis. For one, SCEPTRE did! So quite honestly I don't think this paragraph is appropriate (but see note below). With a heading like "Related work" people might expect to know what is being done in practice.  Your SCEPTRE paragraph is the place to start, but I think we need to expand the lead sentence.  Cite them all and then give a bit about at least one method available in the literature (very very little detail). I'd pick the Monacle based concept and just go with the idea that it is a negative binomial model that uses thresholding and that it sometimes doesn't work and results in a lot of false signals.  Next you can the remainder of your SCEPTRE paragraph, but again I think it makes sense to say something about the basic model from which the CRT is built.  In other words you want to convey that his method still uses thresholds, but solves the false positive problem.  Now you are ready for your final few sentences about how you'll move beyond this. NOTE: Regarding your eSVD paragraph, you could use something like this paragraph in the Discussion, where you could make the point that it is a good thing when models are built on the generating distribution of the data (rather than forcing a normal by transformation).  Then you could say this paper is like the others you cite in doing this.  I make a try below, see the blue paragraph.\\
Here is the alternative idea.  Since my suggestion causes this section to be only one paragraph, you might consider having Assay Overview first and then putting this material at the lead of analysis challenges without a major heading.  If you moved it to there you could do a better job of citing other works for analysis of CRISPR screens.  I am a fan of giving credit to everyone, but doing so with terse sentences that say something like "Several methods have appeared in the literature, each with some shortcomings (ref, ref, ref).".  Anyway, this section could lead people to believe we have very big egos, basically citing ourselves and almost no one else.}
\textcolor{blue}{THIS IS HOW YOU MIGHT REWRITE THE SCEPTRE PARAGRAPH.The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis, each with some shortcomings (cite Gasp, Xie, etc + Barry). The idea is to test whether a given perturbation is associated with the change in expression of a given gene, while ideally adjusting for sources of confounding and ensuring robustness to expression model misspecification. Most relevant is SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}.  SCEPTRE successfully handles the challenges of poor calibration and low power shared by other methods.  In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly,  we account for the fact that the perturbation is measured with noise, an issue that available methods circumvent via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.}
\fi

\subsection{Related work}
Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin et al.\ \parencite{Lin2021} and Townes et al.\ \parencite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions. Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \parencite{Lause2021} and improving performance on dimension reduction tasks. We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on measurement error regression models rather than dimension reduction.

@@

The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis. Most relevantly, Barry et al.\ developed SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.

\subsection{Assay background}
There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \parencite{Gasperini2019,Datlinger2021,Mimitou2019}. In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens, which we motivate and describe here. The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements. Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \parencite{Gallagher2018}. These noncoding variants are thought to contribute to disease by modulating the expression of one or more disease-relevant genes. Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results. A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \parencite{Gasperini2020}.

@@

@@

@@

@@

\section{Thresholding method}
We study thresholding from empirical and theoretical perspectives, highlighting several limitations of the approach. Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence. We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV.

Let $n \in \mathbb{N}$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \mathbb{N}$ be the number of gene transcripts sequenced; let $g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced; let $d^m_i \in \mathbb{N}$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \mathbb{R}^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively. The thresholding method is defined as follows:
\begin{itemize}
	\item[1.] For a given threshold $c \in \mathbb{N}$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be given by $\hat{p}_i = 0$ if  $g_i < c$ and $\hat{p}_i = 1$ otherwise.
	\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
	\begin{equation}\label{thresh_glm}
	m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i); \quad
	\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
	\end{equation}
	where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in  \mathbb{R}^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.

	\item[3.] Fit a GLM to obtain an estimate, CI, and $p$-value for the target of inference $\beta^m_1$.
\end{itemize}
The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression. Exponentiating both sides of (\ref{thresh_glm}) shows that the mean $\mu_i$ of the $i$th cell is equal to
$\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m;$ this quantity is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration. The target of inference $\beta^m_1$ is the log fold change in expression in response to the perturbation, controlling for the technical factors. Fold change in this context is defined as the ratio of the mean gene expression in perturbed cells to the mean gene expression in unperturbed cells. Therefore, $\exp(\beta^m_1) = 1$ (i.e., $\beta^m_1 = 0$) indicates no change in expression, whereas $\exp(\beta^m_1) > 1$ (i.e., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (i.e., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.

\subsection{Empirical challenges of the thresholding method}\label{sec:thresholding_empirical}
To shed light on empirical challenges of the thresholding method, we applied the thresholding method to analyze the set of positive control perturbation-gene pairs in the Gasperini dataset. The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition. Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong decrease in expression. To investigate sensitivity to the choice of threshold, we deployed the thresholding method using three different thresholds: 1, 5, and 20. We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b): estimates for fold change produced by threshold = 1 were smaller in magnitude (i.e., closer to the baseline of $1$) than those produced by threshold = 5. (Figure \ref{thresholding_empirical}a.) On the other hand, estimates produced by threshold = 5 and threshold = 20 were more concordant (Figure \ref{thresholding_empirical}b).

@@

Next, we studied the thresholding method from a theoretical perspective, recovering in a simplified Gaussian setting phenomena revealed in the empirical analysis. Suppose we observe gRNA expression and gene expression data $(g_1, m_1), \dots, (g_n, m_n)$ on $n$ cells from the following linear model:
\begin{equation}\label{theoretical_model}
m_i = \beta^m_0 + \beta^m_1 p_i + \epsilon_i; \quad
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i; \quad
p_i \sim \textrm{Bern}(\pi); \quad
\epsilon_i, \tau_i \sim N(0,1),
\end{equation}
where $p_i, \tau_i,$ and $\epsilon_i$ are independent. For a given threshold $c \in \mathbb{R}$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e. $\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right].$ We derive the almost sure limit of $\hat{\beta}^m_1$: 
\begin{proposition}\label{prop:convergence}
	The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
	\begin{equation}\label{thresh_est_intercepts}
	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \mathbb{E}[ \hat{p}_i ])}{ \mathbb{E}[\hat{p}_i] (1 - \mathbb{E}[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0), 
	\end{equation} where 
	$
	\mathbb{E}[\hat{p}_i] = \zeta(1-\pi) + \omega\pi$, $\omega \equiv \Phi\left(\beta_1^g + \beta_0^g -c \right)$, and $\zeta \equiv \Phi\left( \beta^g_0 - c \right).$
\end{proposition}
The function $\gamma: \mathbb{R}^4 \to \mathbb{R}$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$. The asymptotic relative bias $b: \mathbb{R}^4 \to \mathbb{R}$ of $\hat{\beta}^m_1$ is given by
$$b(\beta^g_1, \pi, c, \beta^g_0)  \equiv \frac{1}{\beta^m_1} \left(\beta^m_1 - \lim_\textrm{a.s.} \hat{\beta}^m_1 \right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0).$$
Having derived an exact expression for the asymptotic relative bias of $\hat{\beta}^m_1$, we can prove several results about this quantity. We fix $\pi$ to $1/2$ for simplicity. (In reality, $\pi$ is smaller, but the relevant statistical phenomena emerge for $\pi = 1/2$.) First, the thresholding estimator strictly \textit{underestimates} (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold $c$ and over all values of the regression coefficients $(\beta^m_0, \beta^m_1)$ and $(\beta^g_0, \beta^g_1)$. This phenomenon, called attenuation bias, is a common attribute of estimators that ignore measurement in errors-in-variables models \parencite{Stefanski2000a}. Second, the magnitude of the bias decreases monotonically in $\beta^g_1$, comporting with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated. Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \mathbb{R}$ (i.e., the most accurate decision boundary for classifying cells) is a critical value of the bias function. Finally, and most subtly, there is no universally applicable rule for selecting a threshold that yields minimal bias: when $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields smaller bias than setting the threshold to the Bayes decision boundary; when $\beta^g_1$ is large, the reverse is true. Appendix \ref{sec:appendix_theory} contains detailed proposition statements and proofs of these results.

Next, we studied the variance of the thresholding estimator, considering a slightly simpler model for this purpose. Suppose the intercepts in (\ref{theoretical_model}) are fixed at $0$ (i.e., $\beta^m_0$ = $\beta^g_0$ = 0). For notational simplicity we write $\beta_m = \beta^m_1$ and $\beta_g = \beta^g_1.$ The thresholding estimator $\hat{\beta}_m$ is the no-intercept OLS solution $\hat{\beta}_m = \left[\sum_{i=1}^n \hat{p}_i^2 \right]^{-1}\left[\sum_{i=1}^n \hat{p}_i m_i \right].$ The following proposition derives the scaled, asymptotic distribution of $\hat{\beta}_m:$
\begin{proposition}\label{prop:bv_decomp}
	The limiting distribution of $\hat{\beta}_m$ is
	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2) }{\left(\mathbb{E}[\hat{p}_i]\right)^2} \right),$$ where $$l \equiv \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
	\mathbb{E}[\hat{p}_i] = \pi \omega + (1-\pi) \zeta; \quad
	\omega \equiv \Phi(\beta_g - c); \quad
	\zeta \equiv \Phi(-c).$$
\end{proposition}
This proposition yields an asymptotically exact bias-variance decomposition for $\hat{\beta}_m$: as the threshold tends to infinity, the bias decreases and the variance increases (Appendix \ref{sec:appendix_theory}). Overall, our empirical and theoretical analyses indicate that thresholded regression poses a basic difficulty: selecting a good threshold is challenging, and even if the threshold is optimal, the estimator could suffer from high bias or high variance. These considerations motivate our research question: does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screens?

@@

To answer this question, we introduce the GLM-EIV model.

>>>>

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in a simplified Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model \parencite{carroll} to responses and noisy predictors that are exponential family-distributed and potentially impacted by the same set of confounding variables. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding. We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.

@@

\iffalse
\subsection{Related work}
Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin et al.\ \parencite{Lin2021} and Townes et al.\ \parencite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions. Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \parencite{Lause2021} and improving performance on dimension reduction tasks. We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on measurement error regression models rather than dimension reduction.

@@

The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis. Most relevantly, Barry et al.\ developed SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.
\fi

\section{Assay background}
There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \parencite{Gasperini2019,Datlinger2021,Mimitou2019}. In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens, which we motivate and describe here. The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of relevance to the current work). Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and inside enhancers \parencite{Gallagher2018}. These noncoding variants are thought to contribute to disease by modulating the expression of one or more disease-relevant genes. Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results. A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \parencite{Gasperini2020}.

@@

The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis \parencite{Gasperini2019,Xie2019a,Barry2021}. Gasperini et al.\ applied negative binomial GLMs (as implemented in the Monocle software; \cite{Trapnell2014}) to carry out the differential expression analysis described above. Xie et al., by contrast, applied chi-squared-like tests of independence for this purpose. Both of these approaches have limitations: the former is not robust to misspecification of the gene expression model, and the latter is unable to correct for the presence of technical confounders. Recently, Barry et al.\ introduced SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}. SCEPTRE simultaneously adjusts for confounder presence and ensures robustness to expression model misspecification, overcoming limitations of the prior methods and demonstrating state-of-the-art sensitivity and specificity on single-cell CRISPR screen data. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we seek to account for the fact that the perturbation is measured with noise, an issue that all available methods (including SCEPTRE) circumvent via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.

@@

\section{Analysis challenges and proposed statistical model}

@@

We propose to model the single-cell CRISPR screen data-generating process using a pair of GLMs. Let $n \in \mathbb{N}$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $p_i \in \{0,1\}$ indicate perturbation presence or absence; let $m_i \in \mathbb{N}$ be the number of gene transcripts sequenced; let $g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced; let $d^m_i \in \mathbb{N}$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); let $d^g_i$ be the gRNA library size; and finally, let $z_i \in \mathbb{R}^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively.

Building on the work of several previous authors \parencite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (\cite{Sarkar2021}). Sarkar and Stephens' framework enjoys strong theoretical and empirical support; therefore, we generalize their approach to model \textit{both} gene and gRNA modalities in single-cell CRISPR screen experiments. To this end, we assume that the gene expression counts are given by
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i^m); \quad \log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where (i) $\textrm{NB}_{\theta^m}(\mu^m_i)$ is a negative binomial distribution with mean $\mu^m_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in  \mathbb{R}^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term. Similarly, we model the gRNA counts by
\begin{equation}\label{glmeiv_model_2}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right); \quad \log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where $\mu^g_i$, $\theta^g$, $\beta^g_0$, $\beta^g_1$, $\gamma_g$, and $d^g_i$ are analogous. We use a negative binomial GLM to model the gRNA counts as well as the gene expressions because the gRNA transcripts are generated via the same biological mechanism as the gene transcripts \parencite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as 
\begin{equation}\label{glmeiv_model_3}
 p_i \sim \textrm{Bern}(\pi); \quad \pi \in (0,1/2],
\end{equation}
where $p_i$ is unobserved. Together, (\ref{glmeiv_model_1} - \ref{glmeiv_model_3}) define the negative binomial GLM-EIV model.

The log-transformed sequencing depth $\log(d^m_i)$ is included as an offset term in (\ref{glmeiv_model_1}) so that $\beta^m_0 + \beta^m_1 p_i + \gamma^T_m z_i$ can be interpreted as a relative expression. Exponentiating both sides of (\ref{glmeiv_model_1}) reveals that the mean gene expression $\mu_i^m$ of the $i$th cell is
$\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$ Because $d^m_i$ is the sequencing depth, $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration. The target of inference $\beta^m_1$ is the log fold change in expression in response to the perturbation, controlling for the technical factors. Fold change in this context is the ratio of the mean gene expression in perturbed cells to the mean gene expression in unperturbed cells. Hence, $\exp(\beta^m_1) = 1$ (i.e., $\beta^m_1 = 0$) indicates no change in expression, whereas $\exp(\beta^m_1) > 1$ (i.e., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (i.e., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.

@@

\section{Analysis of the thresholding method}

We study thresholding from empirical and theoretical perspectives, highlighting several limitations of the approach. In the context of the negative bionomial GLM-EIV model introduced above (\ref{glmeiv_model_1}-\ref{glmeiv_model_3}), thresholding methods use the gRNA counts (\ref{glmeiv_model_2})
to impute the latent perturbation indicator (\ref{glmeiv_model_3}), thereby reducing the full data generating process to a single, gene expression model (\ref{glmeiv_model_1}). We study Gasperini et al.'s  variant of the thresholding method (i.e., thresholded negative binomial regression), as this version of the thresholding method relates most closely to GLM-EIV. The method is defined as follows:

\begin{itemize}
	\item[1.] For a given threshold $c \in \mathbb{N}$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be given by $\hat{p}_i = 0$ if  $g_i < c$ and $\hat{p}_i = 1$ otherwise.
	\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
	\begin{equation}\label{thresh_glm}
	m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu^m_i); \quad
	\log(\mu^m_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right).
	\end{equation}
	The model (\ref{thresh_glm}) is equivalent to the model (\ref{glmeiv_model_2}), but the latent perturbation indicator $p_i$ has been replaced by the imputed perturbation indicator $\hat{p}_i.$

	\item[3.] Fit a GLM to (\ref{thresh_glm}) to obtain an estimate and CI for the target of inference $\beta^m_1$.
\end{itemize}

\subsection{Empirical challenges of the thresholding method}\label{sec:thresholding_empirical}
To shed light on empirical challenges of the thresholding method, we applied thresholded negative binomial regression to analyze the set of positive control perturbation-gene pairs in the Gasperini dataset. The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition. Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong decrease in expression.

To investigate the sensitivity of the thresholding method to threshold choice, we deployed the method using three different choices for the threshold: 1, 5, and 20. We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b): estimates for fold change produced by threshold = 1 were smaller in magnitude (i.e., closer to the baseline of $1$) than those produced by threshold = 5. (Figure \ref{thresholding_empirical}a.) On the other hand, estimates produced by threshold = 5 and threshold = 20 were more concordant (Figure \ref{thresholding_empirical}b).

@@

Next, we studied the thresholding method from a theoretical perspective, recovering in a simplified Gaussian setting phenomena revealed in the empirical analysis. Suppose we observe gRNA expression and gene expression data $(g_1, m_1), \dots, (g_n, m_n)$ on $n$ cells from the following linear model:
\begin{equation}\label{theoretical_model}
m_i = \beta^m_0 + \beta^m_1 p_i + \epsilon_i; \quad
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i; \quad
p_i \sim \textrm{Bern}(\pi); \quad
\epsilon_i, \tau_i \sim N(0,1),
\end{equation}
where $p_i, \tau_i,$ and $\epsilon_i$ are independent. For a given threshold $c \in \mathbb{R}$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e. $\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right].$ We derive the almost sure limit of $\hat{\beta}^m_1$: 
\begin{proposition}\label{prop:convergence}
	The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
	\begin{equation}\label{thresh_est_intercepts}
	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \mathbb{E}[ \hat{p}_i ])}{ \mathbb{E}[\hat{p}_i] (1 - \mathbb{E}[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0), 
	\end{equation} where 
	$
	\mathbb{E}[\hat{p}_i] = \zeta(1-\pi) + \omega\pi$, $\omega \equiv \Phi\left(\beta_1^g + \beta_0^g -c \right)$, and $\zeta \equiv \Phi\left( \beta^g_0 - c \right).$
\end{proposition}
The function $\gamma: \mathbb{R}^4 \to \mathbb{R}$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$. The asymptotic relative bias $b: \mathbb{R}^4 \to \mathbb{R}$ of $\hat{\beta}^m_1$ is given by
$$b(\beta^g_1, \pi, c, \beta^g_0)  \equiv \frac{1}{\beta^m_1} \left(\beta^m_1 - \lim_\textrm{a.s.} \hat{\beta}^m_1 \right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0).$$
Having derived an exact expression for the asymptotic relative bias of $\hat{\beta}^m_1$, we can prove several results about this quantity. We fix $\pi$ to $1/2$ for simplicity. (In reality, $\pi$ is smaller, but the relevant statistical phenomena emerge for $\pi = 1/2$.) First, the thresholding estimator strictly underestimates (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold $c$ and over all values of the regression coefficients $(\beta^m_0, \beta^m_1)$ and $(\beta^g_0, \beta^g_1)$. This phenomenon, called attenuation bias, is a common attribute of estimators that ignore measurement in errors-in-variables models \parencite{Stefanski2000a}. Second, the magnitude of the bias decreases monotonically in $\beta^g_1$, comporting with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated. Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \mathbb{R}$ (i.e., the most accurate decision boundary for classifying cells) is a critical value of the bias function. Finally, and most subtly, there is no universally applicable rule for selecting a threshold that yields minimal bias: when $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields smaller bias than setting the threshold to the Bayes decision boundary; when $\beta^g_1$ is large, the reverse is true. Appendix \ref{sec:appendix_theory} contains detailed proposition statements and proofs of these results.

Next, we studied the variance of the thresholding estimator, considering a slightly simpler model for this purpose. Suppose the intercepts in (\ref{theoretical_model}) are fixed at $0$ (i.e., $\beta^m_0$ = $\beta^g_0$ = 0). For notational simplicity we write $\beta_m = \beta^m_1$ and $\beta_g = \beta^g_1.$ The thresholding estimator $\hat{\beta}_m$ is the no-intercept OLS solution $\hat{\beta}_m = \left[\sum_{i=1}^n \hat{p}_i^2 \right]^{-1}\left[\sum_{i=1}^n \hat{p}_i m_i \right].$ The following proposition derives the scaled, asymptotic distribution of $\hat{\beta}_m:$
\begin{proposition}\label{prop:bv_decomp}
	The limiting distribution of $\hat{\beta}_m$ is
	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2) }{\left(\mathbb{E}[\hat{p}_i]\right)^2} \right),$$ where $$l \equiv \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
	\mathbb{E}[\hat{p}_i] = \pi \omega + (1-\pi) \zeta; \quad
	\omega \equiv \Phi(\beta_g - c); \quad
	\zeta \equiv \Phi(-c).$$
\end{proposition}
This proposition yields an asymptotically exact bias-variance decomposition for $\hat{\beta}_m$: as the threshold tends to infinity, the bias decreases and the variance increases (Appendix \ref{sec:appendix_theory}). Overall, our empirical and theoretical analyses indicate that thresholded regression poses a basic difficulty: selecting a good threshold is challenging, and even if we have selected a threshold that is (in some sense) optimal, the resulting estimator still may suffer from high bias or high variance. These considerations motivate our research question: does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screen analysis?

@@

To answer this question, we introduce the general GLM-EIV model, which generalizes the negative binomial GLM-EIV model (\ref{glmeiv_model_1}-\ref{glmeiv_model_3}) to arbitrary exponential family response distributions and link functions, thereby providing much greater modeling flexibility. We derive efficient methods for estimation and inference in this model and develop a pipeline to deploy the model at-scale. Appendix \ref{sec:zero_inf_model} develops a parallel methodology in which we model gRNA counts as (in some sense) zero-inflated.

\subsection{Model}\label{sec:model}
